---
title: Analysis of U.S. National Oceanic and Atmospheric Administration's (NOAA)
author: "FFadel"
date: "August 14, 2017"
output:
  html_document:
  fig_caption: yes
  html_notebook: default
  pdf_document: default
---
### Sypnosis
This report provide analysis on the type of weather events and the severity of those of harm caused to health and the economy. the NOAA storm data represents the period from 1950 to 2011 where we concluded that tornadoes had the highest fatalities and injuries, while flood caused the largest financial impact (crop and property) through that period.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```
### Data Processing
The data can be directly retrieved using the following [url](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2]). The  script automatically get the data file in bzip (compressed format) and the proceses directly using the `read.csv()` and `bzip()` functions combined which get stored into the `StormData` 
## download files decompress
```{r data, cache=TRUE}
#file url
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

#filename to be read
filename <- "./repdata_data_StormData.csv.bz2"

#check if file already exists otherwise download
if(!file.exists(filename)){
  download.file(url,"./repdata_data_StormData.csv.bz2")
}

#read data into StormData
StormData <- read.csv(bzfile(filename),header=TRUE, stringsAsFactors=FALSE)
```

Verify Data
```{r check data}
#verify data loaded correctly
head(StormData)
```

Cleaning and correcting the features `PROPDMGEXP` and `CROPDMGEXP`, some of the values entered are non numeric and require correction. `B,M,K and H` values  
```{r data cleaning}
unique(StormData$PROPDMGEXP)
unique(StormData$CROPDMGEXP)

#correct values
StormData[StormData$PROPDMGEXP == "K" & StormData$PROPDMGEXP == "k", "PROPDMGEXP"] <- "3"
StormData[StormData$PROPDMGEXP == "M" & StormData$PROPDMGEXP == "m", "PROPDMGEXP"] <- "6"
StormData[StormData$PROPDMGEXP == "B" & StormData$PROPDMGEXP == "b", "PROPDMGEXP"] <- "9"
StormData[StormData$PROPDMGEXP == "H" & StormData$PROPDMGEXP == "h", "PROPDMGEXP"] <- "2"
StormData[!StormData$PROPDMGEXP %in% c(0:9), "PROPDMGEXP"] <- "0"

StormData[StormData$CROPDMGEXP == "K" & StormData$CROPDMGEXP == "k", "CROPDMGEXP"] <- "3"
StormData[StormData$CROPDMGEXP == "M" & StormData$CROPDMGEXP == "m", "CROPDMGEXP"] <- "6"
StormData[StormData$CROPDMGEXP == "B" & StormData$CROPDMGEXP == "b", "CROPDMGEXP"] <- "9"
StormData[StormData$CROPDMGEXP == "H" & StormData$CROPDMGEXP == "h", "CROPDMGEXP"] <- "2"
StormData[!StormData$CROPDMGEXP %in% c(0:9),]$CROPDMGEXP <- "0"
```

### Data Analysis

```{r Storm Fatalities Summary}
StormFatalities <- aggregate(FATALITIES ~ EVTYPE, data=StormData, sum)
names(StormFatalities) <- c("EVTYPE","FATALITIES")
attach(StormFatalities)
StormFatalities[which.max(FATALITIES),]
StormFatalities <- StormFatalities[order(-StormFatalities$FATALITIES),][1:10, ]
StormFatalities$EVTYPE <- factor(StormFatalities$EVTYPE, levels=StormFatalities$EVTYPE)
```

```{r Storm Injuries Summary}
StormInjuries <- aggregate(INJURIES ~ EVTYPE, data=StormData, sum)
names(StormInjuries) <- c("EVTYPE","INJURIES")
attach(StormInjuries)
StormInjuries[which.max(INJURIES),]
StormInjuries <- StormInjuries[order(-StormInjuries$INJURIES),][1:10,]
StormInjuries$EVTYPE <- factor(StormInjuries$EVTYPE, levels=StormInjuries$EVTYPE)
```
### Note: Alphabetical characters used to signify magnitude include "K" for thousands, "M" for millions, and "B" for billions  
```{r Total Damage Calcaulation}
#calculate total damage cost
TOTALDMGCOST <-  StormData$PROPDMG * 10 ** as.numeric(StormData$PROPDMGEXP) +
                 StormData$CROPDMG * 10 ** as.numeric(StormData$CROPDMGEXP)

#append total damage cost to StormData
StormData <- cbind(StormData,TOTALDMGCOST)
```

```{r Storm Damage Summary}
StormDamages <- aggregate(TOTALDMGCOST ~ EVTYPE, data=StormData, sum)
names(StormDamages) <- c("EVTYPE","TOTALDMGCOST")
StormDamages <- StormDamages[order(-StormDamages$TOTALDMGCOST),][1:10,]
StormDamages$EVTYPE <- factor(StormDamages$EVTYPE, levels=StormDamages$EVTYPE)
```

### Results
The respective plots calrify that Tornado (plot 1 & 2) has the highest fatalities and injuries, while flashflooding has more damage impact on properties and crops.
```{r plots}
#plot1 - Top 10 Events by Fatalities
ggplot(StormFatalities, aes(EVTYPE,FATALITIES)) + 
      geom_bar(stat="identity", fill="firebrick") + 
      theme(axis.text.x=element_text(angle=90, hjust=1)) + 
      xlab("Event Type") + 
      ylab("Fatalities") + 
      ggtitle("Top 10 Highest Weather Events with Highest Fatalities")

#plot2 - Top 10 Events by Injuries
ggplot(StormInjuries, aes(EVTYPE,INJURIES)) +
      geom_bar(stat="identity", fill="firebrick4") + 
      theme(axis.text.x=element_text(angle=90, hjust=1)) + 
      xlab("Event Type") + 
      ylab("Injuries") + 
      ggtitle("Top 10 Weather Events with Highest Injuries")

#plo3 - Top 10 Events by Combined Crop & Property Damages
ggplot(StormDamages, aes(EVTYPE,TOTALDMGCOST/1000000)) +
      geom_bar(stat="identity", fill="blue") + 
      theme(axis.text.x=element_text(angle=90, hjust=1)) + 
      xlab("Event Type") + 
      ylab("Damage Cost (Millions)") + 
      ggtitle("Top 10 Weather Events with Highest Damage Cost")
```
